---
title: "VST_counts"
output: html_notebook
---
```{r}
library(ggplot2)
library(DESeq2)
library(tidyverse)
# library(icesTAF)
library(patchwork)
#Load tpm.df
tpm.df <- read.csv("formatted_output/full_merged_tpm.csv",row.names="target_id")
```

#Raw counts data transformations (vsc, log)
Load merged estimated counts table (both int and numeric)
```{r}
merged_counts_int.df <- read.csv("formatted_output/full_merged_count_int.csv",row.names="target_id")
mc_int_matrix <- data.matrix(merged_counts_int.df)
mode(mc_int_matrix) <- "integer"
merged_counts_num.df <- read.csv("formatted_output/full_merged_count.csv",row.names="target_id")
mc_num_matrix = data.matrix(merged_counts_num.df)
```

1. VST of int version of full merged counts
2. log transform of numeric full merged counts 
```{r}
PSEUDOCOUNT = 0.1
transformed_dir = "formatted_output/transformed"
dir.create(file.path(getwd(),transformed_dir),showWarnings = FALSE)
TRANSFORM_FULL_DATA <- FALSE 
if (TRANSFORM_FULL_DATA) {
  vsc.df <- as.data.frame(vst(mc_int_matrix))
  vsc_fpath = sprintf("%s/full_merged_vst.csv",transformed_dir)
  write.csv(vsc.df,vsc_fpath,quote=FALSE)
  
  log_mc = log(mc_num_matrix + PSEUDOCOUNT)
  log_mc.df <- as.data.frame(log_mc)
  log_mc_fpath = sprintf("%s/full_merged_log.csv",transformed_dir)
  write.csv(log_mc.df,log_mc_fpath)
}
```
Plot raw, tpm, vst, and log histograms 
```{r}
PLOT_RAW_HISTS <- FALSE
if (PLOT_RAW_HISTS) {
  tpm_hist <- ggplot(gather(tpm.df, cols, value), aes(x = value)) + 
       geom_histogram(bins=50) 
  tpm_hist + ggtitle("TPM Histogram") + theme(plot.title = element_text(hjust = 0.5))
  
  vst_hist <- ggplot(gather(vsc.df, cols, value), aes(x = value)) + 
         geom_histogram(bins=50) 
  vst_hist + ggtitle("VST Estimated Counts Histogram") + theme(plot.title = element_text(hjust = 0.5))
  
  log_hist <- ggplot(gather(log_mc.df, cols, value), aes(x = value)) + 
         geom_histogram(bins=50) 
  log_hist + ggtitle("Log Counts Histogram") + theme(plot.title = element_text(hjust = 0.5))
}
```
```{r}
#Tpm and prevalence filtering functions#
get_filtered_gene_vec <- function(tpm.df, inp_threshold, inp_prevalence) {
  # tpm.df = locus rows x sample columns
  # inp_threshold = threshold of tpm
  # inp_prevalence = the proportion of samples that are above inp_threshold
  num_samples <- ncol(tpm.df)
  genes_passing <- rownames(data.frame(logic = rowSums(tpm.df > inp_threshold) > num_samples*inp_prevalence) %>% filter(logic == TRUE))
  return(genes_passing)
}
get_prevotella_filtered_gene_vec <- function(tpm.df, inp_threshold, inp_prevalence) {
    #' @param tpm.df locus rows x sample columns
    #' @param inp_threshold threshold of tpm
    #' @param inp_prevalence the proportion of samples that are above inp_threshold
  PREVOTELLA_ROW_RE <- "(NJCFFJJN)|(NBCBLOMG)"
  PREVOTELLA_COL_RE <- "1C"
  num_samples <- ncol(tpm.df)
  prevotella_samples <- grep(PREVOTELLA_COL_RE, colnames(tpm.df))
  prevotella.df <- select(tpm.df,prevotella_samples)
  num_prev_samples <- ncol(prevotella.df)
  #Logic has two cases: 
  #1.number of samples in tpm.df which have tpm > inp_threshold is 
  #greater than number of samples * inp_prevalence
  #2.Prevotella case: number of prevotella containing (1C) samples > inp_threshold is
  #greater than numper of prevotella samples AND row corresponds to prevotella locus
  genes_passing <- rownames(data.frame(logic = ((rowSums(tpm.df > inp_threshold) > num_samples*inp_prevalence) | (rowSums(prevotella.df > inp_threshold) >num_prev_samples*inp_prevalence & (grepl(PREVOTELLA_ROW_RE, row.names(tpm.df)))))) %>% filter(logic == TRUE))
  return(genes_passing)
}
```


```{r}
# inp_threshold = 5 #Steven MAG default
# inp_prevalence = 0.4 #Steven MAG default 
inp_threshold = 3
inp_prevalence = 0.3
prev_filt_genes.vec <- get_prevotella_filtered_gene_vec(tpm.df,inp_threshold,inp_prevalence)
length(prev_filt_genes.vec)
#filt_genes.vec <- get_filtered_gene_vec(tpm.df,inp_threshold,inp_prevalence)
filt_genes.vec <- prev_filt_genes.vec
length(filt_genes.vec)
mc_int.filt.df <- merged_counts_int.df[rownames(merged_counts_int.df) %in% filt_genes.vec,]
mc_num.filt.df <- merged_counts_num.df[rownames(merged_counts_num.df) %in% filt_genes.vec,]
filt_tpm.df <- tpm.df[rownames(tpm.df) %in% filt_genes.vec,]
#vst and log of filtered dfs
mc_filt_matrix <- data.matrix(mc_int.filt.df)
mode(mc_filt_matrix) <- "integer"
filt_vst.df <- as.data.frame(vst(mc_filt_matrix)) #parametric fitType fails and forces local regression
filt_rlog.df <-as.data.frame(rlog(mc_filt_matrix))
filt_log.df <- as.data.frame(log(data.matrix(mc_num.filt.df) + PSEUDOCOUNT))
#Write filtered dataframes to csv 
filt_vst_fpath = sprintf("%s/filt_merged_vst.csv",transformed_dir)
write.csv(filt_vst.df,filt_vst_fpath,quote=FALSE)
filt_rlog_fpath = sprintf("%s/filt_merged_rlog.csv",transformed_dir)
write.csv(filt_rlog.df,filt_rlog_fpath,quote=FALSE)
filt_log_fpath = sprintf("%s/filt_merged_log.csv",transformed_dir)
write.csv(filt_log.df,filt_log_fpath,quote=FALSE)
filt_tpm_fpath = sprintf("%s/filt_merged_tpm.csv",transformed_dir)
write.csv(filt_tpm.df,filt_tpm_fpath,quote=FALSE)
```

```{r}
#Histogram plotting 
#TODO: superimpose prevotella 2A samples (expected zeroes) on histograms
#' Return subset of data for Prevotella mapped transcripts from samples without colonization.
#' @param expr.df DataFrame containing expression data. 
#' @param PREVOTELLA_ROW_RE String for regular expression corresponding that matches all Prevotella loci 
#' @param NON_PREVOTELLA_COL_RE regular expression string matching all samples lacking prevotella
get_prevotella_2A_data <- function(expr.df,PREVOTELLA_ROW_RE= "(NJCFFJJN)|(NBCBLOMG)",NON_PREVOTELLA_COL_RE="2A") {
  prev_rows.df <- filter(expr.df,grepl(PREVOTELLA_ROW_RE,rownames(expr.df)))
  prev_rows_np_samples <- select(prev_rows.df,grep(NON_PREVOTELLA_COL_RE, colnames(prev_rows.df)))
  return(prev_rows_np_samples)
}
#' Flatten expr.df into two vectors, corresponding to samples with and without expected expression
#'@param expr.df DataFrame containing expression data per locus; must have column names with and without NON_PREVOTELLA_COL_RE
#'@param PREVOTELLA_ROW_RE regular expression which matches rows corresponding to loci from either Prevotella stercorea or Prevotella copri  
#'@param NON_PREVOTELLA_ROW_RE regular expression matching columns corresponding to samples which were not gavaged with Prevotella sp 
#'Returns: output, list containing 1. expected_expr_samples and 2. expected_no_expr_samples
#'1. expected expr_samples: vector of expression data for both non-prevotella loci and prevotella loci from prevotella-colonized samples 
#'2. expected_no_expr_samples: vector of expression data for prevotella loci from samples not colonized with Prevotella (ie "expected zeroes")
vector_prev2A_split <- function(expr.df,split="expected",PREVOTELLA_ROW_RE= "(NJCFFJJN)|(NBCBLOMG)",NON_PREVOTELLA_COL_RE="2A") {
  prev_rows.df <- filter(expr.df,grepl(PREVOTELLA_ROW_RE,rownames(expr.df)))
  non_prev_rows.df <- filter(expr.df,!grepl(PREVOTELLA_ROW_RE,rownames(expr.df)))
  prev_rows_np_samples.df <- select(prev_rows.df,grep(NON_PREVOTELLA_COL_RE, colnames(prev_rows.df)))
  prev_rows_p_samples.df <- select(prev_rows.df,!grep(NON_PREVOTELLA_COL_RE, colnames(prev_rows.df)))
  #Flatten DataFrames into vectors
  non_prev_rows <- unlist(non_prev_rows.df)
  prev_rows_p_samples <- unlist(prev_rows_p_samples.df)
  prev_rows_np_samples <- unlist(prev_rows_np_samples.df)
  #Combine non_prev_rows and prev_rows_p_samples int expected expr_samples
  expected_expr_samples <- c(non_prev_rows,prev_rows_p_samples)
  expected_no_expr_samples <- prev_rows_np_samples
  if (split=="expected") {
    split_samples <- list(expected_expr_samples,expected_no_expr_samples)
  }
  else if (split=="prevotella") {
    split_samples <- list(non_prev_rows, prev_rows_p_samples, prev_rows_np_samples)
  }
  return(split_samples)
}
#' Generate a histogram of expression data values split by either prevotella/non-prevotella loci or by loci expected/not-expected expression based on colonization status 
#' @param expr.df DataFrame containing expression data per locus to be plotted
#' @param split_type "expected" or "prevotella" corresponding to split options in above descrtiption; "expected" - split data into two portions, expected and not-expected expression based on colonization; "prevotella" - three portions: non-prevotella loci/prevotella loci in colonized samples/ prevotella loci in non-colonized samples 
#' @return Hstogram ggplot object 
#' 
expr_histogram <- function(expr.df,expr_type="log",split_type="expected",PREVOTELLA_ROW_RE= "(NJCFFJJN)|(NBCBLOMG)",NON_PREVOTELLA_COL_RE="2A") {
  #Generate corresponding split of data either into two vectors by expected/non-expected expression or into three vectors by non-prevotella loci/prevotella loci in colonized samples/ prevotella loci in non-colonized samples 
  split_samples <- vector_prev2A_split(expr.df,split=split_type,PREVOTELLA_ROW_RE=PREVOTELLA_ROW_RE,NON_PREVOTELLA_COL_RE=NON_PREVOTELLA_COL_RE)
  if (split_type=="expected") {
    expected_expr <- split_samples[[1]]
    expected_no_expr <- split_samples[[2]]
    split.df <- data.frame(variable=c(rep("Colonized Loci",length(expected_expr)),
                                      rep("Non-colonized loci",length(expected_no_expr))),
                           value=c(expected_expr,expected_no_expr))
  }
  else if (split_type == "prevotella") {
    np_loci <- split_samples[[1]]
    p_loci_p_samples <- split_samples[[2]]
    p_loci_np_samples <- split_samples[[3]]
    split.df <- data.frame(variable=c(rep("Non-Prevotella Loci",length(np_loci)),
                                      rep("Prevotella loci, Prevotella samples",length(p_loci_p_samples)),
                                      rep("Prevotella loci, Non-Prevotella samples",length(p_loci_np_samples))),
                           value=c(np_loci,p_loci_p_samples,p_loci_np_samples))
  }
  #Histogram Plotting of split data; position = identity for non-stacked histogram
  hist <- ggplot(split.df,aes(x=value,fill=variable))+geom_histogram(bins=50,alpha=0.7,position="identity")
  hist_title <- sprintf("Filtered %s Estimated Counts",expr_type)
  hist <- hist + labs(title=hist_title,x="Expression",y="Counts") + theme(plot.title = element_text(hjust = 0.5))
  hist <- hist + guides(shape=guide_legend(override.aes = list(size = 0.5)))
  hist <- hist + guides(color=guide_legend(override.aes = list(size = 0.5)))
  hist <- hist + theme(legend.title = element_text(size = 6), 
               legend.text = element_text(size = 6))
  return(hist)
}
#Expression Hists for vst, log, rlog - Expected expression based on colonization split 
vst_hist_expected_split <- expr_histogram(filt_vst.df,expr_type="vst",split_type="expected")
log_hist_expected_split <- expr_histogram(filt_log.df,expr_type="log",split_type="expected")
rlog_hist_expected_split <- expr_histogram(filt_rlog.df,expr_type="rlog",split_type="expected")
# #Patchwork combine - expected split
# expected_combined_hist <- vst_hist_expected_split + log_hist_expected_split + rlog_hist_expected_split + plot_layout(ncol = 3, guides = "collect")
#Expression Hists for vst, log, rlog - Non-Prevotella/Prevotella loci split
vst_hist_prev_split <- expr_histogram(filt_vst.df,expr_type="vst",split_type="prevotella")
log_hist_prev_split <- expr_histogram(filt_log.df,expr_type="log",split_type="prevotella")
rlog_hist_prev_split <- expr_histogram(filt_rlog.df,expr_type="rlog",split_type="prevotella")
# #Patchwork combine - prevotella split 
# prev_combined_hist <- vst_hist_prev_split + log_hist_prev_split + rlog_hist_prev_split + plot_layout(ncol = 3, guides = "collect")
#Align y-axis for all distributions within a split type 
SYNC_YLIM <- TRUE
if (SYNC_YLIM) {
  #Patchwork combine - expected splits, prevotella splits
  combined_exp_prev <- vst_hist_expected_split + log_hist_expected_split + rlog_hist_expected_split +
  vst_hist_prev_split + log_hist_prev_split + rlog_hist_prev_split +plot_layout(ncol = 3, guides = "collect") #(2,3) expected hist / prevotella hist
  
  exp_ranges_y <- c(ggplot_build(combined_exp_prev[[1]])$layout$panel_scales_y[[1]]$range$range,ggplot_build(combined_exp_prev[[2]])$layout$panel_scales_y[[1]]$range$range, ggplot_build(combined_exp_prev[[3]])$layout$panel_scales_y[[1]]$range$range)
  prev_ranges_y <- c(ggplot_build(combined_exp_prev[[4]])$layout$panel_scales_y[[1]]$range$range,ggplot_build(combined_exp_prev[[5]])$layout$panel_scales_y[[1]]$range$range,ggplot_build(combined_exp_prev[[6]])$layout$panel_scales_y[[1]]$range$range)
  for (i in 1:3) {
    combined_exp_prev[[i]] <- combined_exp_prev[[i]] + ylim(min(exp_ranges_y),max(exp_ranges_y))
    combined_exp_prev[[i+3]] <- combined_exp_prev[[i+3]] + ylim(min(prev_ranges_y),max(prev_ranges_y))
  }
} else {
  combined_exp_prev <- vst_hist_expected_split + log_hist_expected_split + rlog_hist_expected_split + vst_hist_prev_split + log_hist_prev_split + rlog_hist_prev_split +plot_layout(ncol = 3, guides = "collect") #(2,3) expected hist / prevotella hist
}
#Save figure 
hist_fpath <- sprintf("figures/expression_hist/tpm%s_prev%s_transformed_hist.png",inp_threshold,inp_prevalence)
print(combined_exp_prev)
ggsave(hist_fpath,width=12,heigh=8,units="in")
```


#Testing for qqplots
```{r}
for (expr.df in list(filt_vst.df,filt_log.df,filt_rlog.df)) {
  # prevotella_split_samples <-vector_prev2A_split(filt_rlog.df,split="prevotella")
  prevotella_split_samples <-vector_prev2A_split(expr.df,split="prevotella")
  np_loci <- prevotella_split_samples[[1]]
  expr_vec <- np_loci
  par(mfrow=c(1,1))
  DEFAULT_TITLE="Non-prevotella loci"
  expr_type <- "rlog"
  title_str <- sprintf("%s %s expression Q-Q Plot",DEFAULT_TITLE,expr_type)
  # qqp <- qqnorm(expr_vec,main=title_str) 
  # qqline(expr_vec)
  ks_test <- ks.test(expr_vec,'pnorm',exact=TRUE)
  ks_p <- ks_test$p.value
  ks_D <- ks_test$statistic
  print(ks_p)
  print(ks_D)
}
```



TODO: Test different tpm and prevalence thresholds (paramter sweep), assess normality and 
number of genes retained 
TODO: Compare vst and log of tpm vs vst/log of counts